# Lets do some math!

```{r echo = FALSE, eval = TRUE, message = FALSE}
library('ggplot2')
library('dplyr')
library('deSolve')
library('tidyr')
```

## Models

***
\begin{center}
\begingroup\Large
Simple statistical model  
\endgroup
\end{center}
***

OK. There is no such thing as simple statistical model. However there are lots of packages that will make you suffer less. In fact this is one of biggest **R** advantages, that you can make even very sophisticated statistical modelling without any knowledge on programming since you use *black boxes*. When you are dealing with classic statistical models many of them are included in **base R** distribution -- like linear or generalized additive models. You probably need to use mixed effect models, at some point. Good news is that there is a very nice and quite straightforward to use package called `lme4`. Every time you run into problem, and you do not know what to use for statistical modelling, or how to perform full procedure, just query Google. There are hundreds of blogs, web pages and *Stack Overflow* discussion on it.

***
\begin{center}
\begingroup\Large*
Other models  
\endgroup
\end{center}
***

Besides *simple statistical models*, there is a lot of other models. The scope of this book is not to discuss dichotomies in classification of models. To keep things simple, we will just stick to *models* that can be described with at least one of the following adjectives: *stochastic*, *mechanistic* or belong to *differential equations systems*. 

## Packages
Other, usually dynamic models, require more knowledge, experience and some packages. Before you start looking for more tailored solutions, install following packages: `deSolve`, `fitdistrplus`, `rriskDistributions` and `truncdist`. First one contains tools for solving differential equations sets, second and third provides you tools to deal with distributions (such as comparing distributions or estimating its parameters), and the last one allows you to use truncated distributions.

## Simple mechanistic model and noise
To show some basic mechanistic model, we will use the well known Michaelis-Menten function: $f(x) = \frac{ax}{b+x}$, with *parameter a = 2.15* and *parameter b = 0.08*.

```{r MMfunction, fig.height = 4, fig.cap = 'Simple Michaelis-Menten function', fig.show = 'hold'}
parA <- 2.15
parB <- 0.08
varX <- seq(0,1, length.out = 1000)
mmFunction <- parA * varX/(parB + varX)
plot(mmFunction~varX, type = 'l')
```

As you can see, as long as you have simple equation it is very straightforward to translate it into R. But we want something more useful, for instance adding noise or uncertainty to our model. Lets assume, that we are not sure what is the value of *parameter a*. Lets say that from previous research and expert knowledge we assume that it can be as high 2.5, but never is lower than 1.8. Also it usually takes value of 2.15. Knowing this we can use PERT distribution to include uncertainty in model.

```{r message = FALSE}
library('mc2d')
sParA <- rpert(1000, 1.8, 2.15, 2.5)
parB <- 0.08
mmSAFunction <- sParA * varX/(parB + varX)
```

We added some uncertainty to our function. But what with *parameter b*, how our uncertainty on its value affects the outcome? Lets assume that it's value is normally distributed with *mean = 0.8* and *sd = 0.23*.

```{r fig.height = 4, fig.show = 'hold'}
sParB <- rnorm(1000, 0.08, 0.023) 
mmSABFunction <- sParA * varX/(sParB + varX)
```

Let me put this two functions on plots side by side:

```{r MMfunctionAB, fig.height = 4, fig.cap = 'Michaelis-Menten function with stochastic parameter A (left) or A and B (right)', fig.show = 'hold'}
par(mfrow = c(1,2))
plot(mmSAFunction~varX, type = 'l')
plot(mmSABFunction~varX, type = 'l')
par(mfrow = c(1,1))
```

It seems that inclusion of uncertainty on value of *parameter b* does change our outcome. Lets also include some variability into the outcome. We can assume that for some reasons value of function will vary more than just our uncertainty about value of both *parameters*. And usually it varies from $f(x)$ by some value from uniform distribution. The 25^th^ and 75^th^ percentile are given by: `-0.13` and `0.17`. To calculate parameters of this distribution we will use `rriskDistributions` package.

```{r fig.height = 4, message = FALSE, fig.cap = 'Michaelis-Menten function fully stochastic', , fig.show = 'hold'}
library('rriskDistributions')
unifParams <- get.unif.par(p = c(0.25, 0.75), q = c(-0.13, 0.17), plot = F)
fVariability <- runif(1000, min = unifParams[1], max = unifParams[2])
mmFSFunction <- ((sParA * varX/(sParB + varX)) + fVariability)
plot(mmFSFunction~varX, type = 'l')
```

It seems that there are differences between those three plots. So lets overlay them one on another to inspect visually differences.

```{r fig.height = 4, fig.cap = 'Michaelis-Menten function with different levels of stochasticity', fig.show = 'hold'}
plot(varX, mmFunction, type = 'l',
     col = 'black', ylim = c(0, 3),
     main = 'Michaelis-Menten function', xlab = 'x', ylab = 'f(x) = a*x/(b+x)')
lines(varX, mmSAFunction, col = 'blue')
lines(varX, mmSABFunction, col = 'green', lty = 'dashed')
lines(varX, mmFSFunction, col = 'red', lty = 'dotted')
```

## Solving differential equations

### The simple epidemiological
SIR (Susceptible, Infectious, Recovered) is a typical compartmental model for epidemiology. Since it has only 3 compartments, it is hard to find an easier one to model with differential equations. In our example we will use slightly more complicated example with four compartments -- SEIR (Susceptible, Exposed, Infectious, Recovered). We begin with defining parameters of initial population. We need to define: birth and death rate, transmission ($\beta$), $\gamma$ ($\frac{1}{\gamma}$ defines infectious period) and $\alpha$ ($\frac{1}{\alpha}$ defines latent period). Than we define time and initial conditions -- number of all specimens, population size, number of exposed, number of initially infected, number of initially recovered, which we combine in `initial` values vector.  
Main step of this procedure is to define function which will be used to solve *Ordinary Differential Equations (ODE)*. This function takes three parameter -- t (which is time sequence), state (which are our initial conditions) and parameters. In the body of the function we define differential equations which connect our compartments. To solve the system of differential equations we use `ode()` function from `deSolve` package and save our results in `modelOutput` variable, which we will later use to make a plot.

```{r}
parameters <- c(b     = 0.1,        # birth rate
                d     = 0.1,        # death rate
                beta  = 0.00025,    # transmission parameter
                gamma = 0.6,        # 1/gamma = infectious period
                a     = 0.1        # 1/a = latent period
                )

time <- 1:100

# Initial conditions
N = 15000    # population size
E0 = 150     # initialy exposed
I0 = 15      # initialy infected
R0 = 20      # initialy recovered

initial <- c(S = N - (E0 + I0 + R0), E = E0, I= I0, R = R0)

# ODE Model
fSEIR <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS = b * (S + E + I + R) - beta * S * I - d * S
    dE = beta * S * I - a * E - d * E
    dI = a * E - gamma * I - d * I
    dR = gamma * I - d * R
    return(list(c(dS, dE, dI, dR)))
  })
}

modelOutput <- deSolve::ode(y = initial, func = fSEIR, times = time, parms = parameters)
head(modelOutput)
```

### Ploting SERI model
To create a plot using our favorite library `ggplot2` (which will be described with more details in chapter \@ref(graphs)) we first need to convert our modelOutput to object of class *data frame*. Than we can use code below to plot our model.

```{r}
modelOutput <- as.data.frame(modelOutput)
ggplot(modelOutput, aes(x = time)) +
  geom_line(aes(y = S, colour = "S")) + 
  geom_line(aes(y = E, colour = "E")) +
  geom_line(aes(y = I, colour = "I")) + 
  geom_line(aes(y = R, colour = "R"))
```

However, the element `geom_line` is repeated four times, which does not look nice. We can change it with little effort, and also define some better aesthetics:

```{r}
modelOutput <- as.data.frame(modelOutput) %>%
  gather(key = compartment, value = value, -time) %>%
  left_join(data.frame(compartment = c("S", "E", "I", "R"),
                       ordered = 1:4,
                       stringsAsFactors = F), by = 'compartment')
ggplot(modelOutput, aes(x = time, y = value,
                        colour = reorder(compartment, ordered))) +
  geom_line() +
  viridis::scale_colour_viridis(discrete = T) +
  theme_bw() +
  theme(panel.grid.major.y = element_line(colour = "#dfdfdf")) +
  labs(colour = 'Compartment',
       x = 'Time',
       y = 'Population size',
       title = 'SERI epidemiological model')
```

You might wonder why we added additional column to `modelOutput` data frame. The thing is that `ggplot` automatically maps some aesthetics to groups using alphabetical order. Created column in which we store numbers that refer to compartment order in *SERI* acronym. Than we use it as argument to `reorder` function inside `ggplot`, which under the hood transforms variable to *factor* and change *level* orders. Using this approach, we don't need to change all the orderings of variables and aesthetics manually. I also used color palette from *`viridis` package, which contains four nice looking color palettes, which are colorblind safe (at least in theory).
