# Functions

In short, function is a piece of code that takes some arguments, makes *mumbo jumbo* and returns result. All the time, through this book we were using functions that are buil in **base R** or comes from additional packages (like `dplyr`). So you are now quite familiar whit the syntax that resambles typical mathematical syntax -- *name of a function followed by arguments in brackets - like f(x)*. From time to time you will need to do something in your code for few times with different arguments. In order to not repeat yourself and not *copy -- paste* your code multiple times you can wrap your procedure in a function. The best way (as always) to understand how to do it, is to do it. To begin with something simple we will start with makeing basic math functions which later will lead us to simple calculator.

## Simple math functions.
Simple math operations include: *adding*, *substracting*, *dividing* and *multipling*. To make our calculator slightly less boring, we can also add *powers* and *nth rooting*. To not to complicate to much things in the begining, lets say that we want our function to take two and only two arguments. Lets look below how to code our functions:

```{r echo = TRUE, eval = FALSE}
addF <- function(x,y) {
  return(x+y)
}
subF <- function(x,y) {
  return(x-y)
}
divF <- function(x,y) {
  return(x/y)
}
mulF <- function(x,y) {
  return(x*y)
}
powF <- function(x,y) {
  return(x:y)
}
ntrF <- function(x,y) {
  return(x^(1/y))
}
```

## Building your own calculator
Above example is of course useless and boring. So lets quickly get to making calculator, that would use one of above functions, or return results of all of them. So actually this time we will make so called *wraper* around previously made functions. We will use *switch* functionality, so besides our two parameters: `x` and `y` we will also tell our function which of the results we are interested in.

```{r}
simpleCalculator <- function(x, y, mathType) {
  addF <- function(x, y) {
  return(x+y)
  }
  subF <- function(x, y) {
  return(x-y)
  }
  divF <- function(x, y) {
  return(x/y)
  }
  mulF <- function(x, y) {
  return(x*y)
  }
  powF <- function(x, y) {
  return(x^y)
  }
  ntrF <- function(x, y) {
  return(x^(1/y))
  }
  switch(mathType,
         add = addF(x, y),
         substract = subF(x, y),
         multiple = mulF(x, y),
         divide = divF(x, y),
         power = powF(x, y),
         root = ntrF(x, y),
         all = cat('The result of mathematical opereators on two numbers:',
                      paste(x, 'and', y), 'are:', '\naddition:', addF(x, y),
                      '\nsubstraction:', subF(x, y), '\nmultiplication:', mulF(x, y),
                      '\ndvision:', divF(x, y), '\nWhat is more the', y,
                      'th power of', x, 'is', powF(x, y), 'and', x, y,
                      'th root is', ntrF(x, y)))
}

simpleCalculator(25,5, 'root')
simpleCalculator(16,2,'all')
```

## It is not over yet... *Calculator shouldn't divide be 0*!

We know that division by 0 is not the best idea in the world, thus we should stop users (or ourselves) from doing it. Thus we will add an `if...else` statment to our function. Next time when you use 0 as a second argument you will see an error. Also, we declare `mathType = 'all'` as a default value, so if we omit this parameter, function will evaluate anyway.

```{r}
simpleCalculator <- function(x, y, mathType = 'all') {
  if (mathType == 'divide' & y == 0) {
    return('You cannot divide by 0, please change y value.')
  } else if (mathType == 'root' & y == 0) {
    return('Root denominator is 0, cannot perform operation, please change y value.')
  } else if (mathType == 'all' & y == 0) {
    return('Y value needs to be different from 0 to make division and nth root.')
  }
  addF <- function(x, y) {
  return(x+y)
  }
  subF <- function(x, y) {
  return(x-y)
  }
  divF <- function(x, y) {
  return(x/y)
  }
  mulF <- function(x, y) {
  return(x*y)
  }
  powF <- function(x, y) {
  return(x^y)
  }
  ntrF <- function(x, y) {
  return(x^(1/y))
  }
  switch(mathType,
         add = addF(x, y),
         substract = subF(x, y),
         multiple = mulF(x, y),
         divide = divF(x, y),
         power = powF(x, y),
         root = ntrF(x, y),
         all = cat('The result of mathematical opereators on two numbers:',
                      paste(x, 'and', y), 'are:', '\naddition:', addF(x, y),
                      '\nsubstraction:', subF(x, y), '\nmultiplication:', mulF(x, y),
                      '\ndvision:', divF(x, y), '\nWhat is more the', y,
                      'th power of', x, 'is', powF(x, y), 'and', x, y,
                      'th root is', ntrF(x, y)))

}
simpleCalculator(25,0)
simpleCalculator(25,0, 'root')
simpleCalculator(25,0, 'divide')
simpleCalculator(25,5)
```