# Graphics

```{r echo = FALSE, eval = TRUE, message = FALSE}
library('ggplot2')
library('dplyr')
```

What would be our work value without visualization? Not much. **R** provides use with some tools to make plots, charts and other visual stuff. However base version has very limited graphic design by default and making it pretty needs a lot of time and code. Nowadays, however, in `tidyverse` there is a very powerful library with dozens of extensions - `ggplot2`. *GG* stands for *Gramar of Graphics* and in practice it means that we build our visualization layer after layer. The vast spectrum of `ggplot` functions and its extensions is out of the scope of this book. Here we will focus on the basic and most useful things as well as how to find not so common functionalities.

## Base plots
Before we get to `ggplot2`, we should begin with base graphic functions. It is true that the do not look as pretty as plots from dedicated libraries, but they are extremely useful in quick checking of our workflow. Not to mention, that many packages are still operating with old fashioned base graphics.

## How to make plot?
In previous chapters we made some plots, but we did not cover the details. Most important thing is to acknowledge that we can make plots with `hist()`, `plot()`, `barplot()` and `boxplot()` functions (to mention only the most common). There are of course other types of plots you might need, however those three are the *classic* ones. If you are looking for more examples of base plots, you should visit help page of `graphics` package. To change the look of your plot, just provide proper arguments to this function, like `col` for color of line/points or, `xlab` and `ylab` to change names of axis labels. Unfortunately a list of arguments that you can control and adjust is long. The good news is that all of them have their default value, so you need to change only the ones you want, without thinking about rest. Full list of parameter you can change you will find after executing: `?par` command. As this help page contains many different kinds of parameters you should scroll to find *Graphical Parameters* chapter.

## Can we actually do something?
Yes we can. But as I said earlier scope of this book is not graphics. Thus I will cover here only the basic things you should know, or at least know where to find them. Bellow, I will present only code and its output. After what you read, you should be perfectly fine with figuring out what is going on. You can also copy - paste the code in your script or console, edit it and learn how it works by doing. Even the best book is just a book, and nothing will substitute the knowledge you gain by experiencing.  

```{r fig.height = 4.5}
graphicDummy <- data.frame(x = rnorm(100, 10, 2), y = rpois(100, 1.2))
par(mfrow = c(2,4))
plot(graphicDummy$x~graphicDummy$y)
plot(graphicDummy$x~as.factor(graphicDummy$y))
boxplot(graphicDummy$x~graphicDummy$y, main = 'Boxplot')
barplot(graphicDummy$x, main = 'Barplot')
hist(graphicDummy$x, main = 'Histogram x')
hist(graphicDummy$y, main = 'Histogram y')
plot(density(graphicDummy$x), main = 'Density x')
plot(density(graphicDummy$y), main = 'Density y')
```

## `ggplot2` - Graphics taken into other dimension
This package use different approach to make plots and charts (however it is not limited to only those). The philosophy it follows is called Grammar of Graphics [@wilkinson2005; @layered-grammar]. Within this workframe, we build whole plot (chart, etc.) by adding layer after layer of visual options. *With great power comes great responsibility* -- `ggplot2` as well as it's extensions are highly customisable, thus if you want to use non-default settings it may take some time before you figure out how to do it. Here we will briefly cover the most basic stuff, but there is very good book that goes deeper into manipulation of visual effects, with lots of examples -- R Graphics Cookbook [@chang2013] which is also available [on line](http://www.cookbook-r.com/Graphs/).

*Basic plot*
Making basic plot requires defining 2 layers:

1. Data
2. Type of plot/chart

```{r fig.height = 2.5}
plotTable <- data.frame(time = 1:10,
                        depVal = c((1:10)*2)+2,
                        typeDep = rep(c('first', 'second'), times = 5))
basePlot <- ggplot(plotTable, aes(x = time, y = depVal)) +
  geom_point(aes(colour = typeDep, alpha = 0.8, shape = typeDep, size = 3.5),
             show.legend = FALSE)
basePlot
```

In the function call (1^st layer), we define data set (`plotTable`) and aesthetics (`x` and `y` values). In following layer, we define 3 aesthetics -- `colour`, `alpha`, `shape` of points and their `size`. We also tell `ggplot` to not to make legend. Actually we obtain the same plot using a little bit different syntax. Unfortunately, when you first start working with `ggplot` (and even when you have some experience), you will often run into small problems that will need some adjustments in your code. Hopefully, this package is so broadly used that you will easily find a lot of solution of your problems on *Stack Overflow* as well as many extensions with predefined functions to make sophisticated graphics.

Lets take a little deeper look into layers and aesthetics. Default look of plot is little bit boring, and usually unsuitable for publications. In code below we will use another layers to define plot background, plot and axis names.

```{r fig.height = 2.5}
basePlot +
  labs(title = 'Basic dot plot', x = 'Time', y = 'Dependent variable') +
  theme(axis.title.x = element_text(size = rel(0.7), colour = 'red'),
        axis.title.y = element_text(size = rel(1.1)), 
        plot.background = element_rect(fill = 'yellow'),
        panel.background = element_rect(fill = 'pink'),
        panel.grid = element_blank())
  
```
Beautiful!  
However we can use simple trick to make our plot in format that is actually supported by scientific journals:
```{r fig.height = 2.5}
basePlot +
  labs(title = 'Basic dot plot', x = 'Time', y = 'Dependent variable') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

*Density plots*
But what if we want to actually plot something useful? Lets say that we want to graphically compare distributions of two data sets. We can do it by histograms (boring) or by plotting density functions. First we should generate some data. Assume that you are tossing coin, but you have this strange gut feeling that something is not right. In 500 trials you had 218 tails and 282 heads.

```{r fig.height = 2.5}
coinDensity <- data.frame(coin = c(rep(1, times = 282), rep(0, times = 218)))
ggplot(coinDensity, aes(x = coin)) + geom_density()
```

And we want to check if it is really different from binomial distribution.
```{r fig.height = 2.5}
tCoinDensity <- data.frame(coin = rbinom(500, 1, 0.5))
ggplot(tCoinDensity, aes(x = coin)) + geom_density()
```

Now we have two plots which is not the best way to visually check two distributions. It would be much better if we plot them one by one, preferably with different colors and opacity. Below it's short code snippet how to do it in **R**.

```{r fig.height = 2.5}
coinTest <- bind_rows(coinDensity, tCoinDensity) %>%
  bind_cols(distribution = rep(c('empirical', 'theoretical'), each = 500))
ggplot(coinTest, aes(x = coin, fill = distribution)) + 
  geom_density(alpha = 0.7, colour = 'yellow') +
  scale_fill_manual(values = c('blue', 'violet'))

```

Indeed, the coin is slightly biased towards heads.

*Box-plot*
When dealing with comparison of two or more groups we usually check their distribution by visually inspecting so called *box plots* or *box-whiskers plots*. This plot shows interquartile range (IQR) as box, median as a line inside box, 1.5 IQR as *whiskers* and all of the data points that are more than 1^st quartile - 1.5 IQR or 3^rd quartile + 1.5 IRQ as separated dots called *outliers*. There are few different approaches to define *outlieres* and *whiskers* in box plots, nonetheless above is most common. It is very straightforward to make box-plots with `ggplot2`:

```{r fig.height = 2.5}
boxGroups <- data.frame(group = rep(c('normal', 'chisq', 'pois'), each = 100),
                        values = c(rnorm(100, 3, 0.65), rchisq(100, 3), rpois(100, 3)))
ggplot(boxGroups, aes(x = group, y = values)) +
  geom_boxplot()
```

*Bar plot*
This visualization technique is mainly used to show categorical variables as rectangles with height (length) that is proportional to value that each categorical variable represents. It is also very easy to create bar-plots in `ggplot2`:

```{r fig.height = 2.5}
barGroups <- boxGroups %>% group_by(group) %>% summarise(meanVal = mean(values))
ggplot(barGroups, aes(x = group, y = meanVal)) +
  geom_bar(stat = 'identity')
```


*Regression*
Last common visualization is regression plot. With layer by layer approach we can control all the elements of the plot. Below code shows how you can visualize both dots and lines for regression, change title and labels and set some non standard appearance of plot.

```{r fig.height = 3}
regVar <- c(sample(1:20, 100, replace = T))
regRes <- ((regVar * 2.6) + 0.8) + order(rnorm(100, 10, 4.25))
regDF <- data.frame(x = regVar, y = regRes)
ggplot(regDF, aes(x, y)) +
  geom_point(fill = 'yellow', colour = '#ff0099', shape = 21, stroke = 1.15) +
  geom_ribbon(stat='smooth', method = "lm", se = TRUE, 
              fill = '#990011', colour = 'yellow', alpha = 0.4,
              linetype = 3, size = 0.2) +
  geom_line(stat='smooth', method = "lm",
            colour = 'green', alpha = 1,
            linetype = 4, size = 1.15) +
  theme(panel.background = element_rect(fill = NA),
        axis.line = element_line(size = 1.15, colour = 'grey')) +
  labs(x = 'Values of X',
       y = 'y = 2.6 * x + 0.8',
       title = 'Example of regression plot') +
  annotate('text', label = 'y==2.6*x+0.8+epsilon', parse = T,
           x = 3.5, y = 135, colour = 'blue')
```
