---
title: "Examples from Beginners Guide to R Galaxy"
author: "Michal Czyz"
date: "May 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 2

Vectors
```{r echo = T, eval = T}
characterVector <- c('a', 'b', 'c')
class(characterVector)
integerVector <- c(1L, 2L, 3L)
class(integerVector)
numericVector <- c(2.5, 3.5, 4.5)
class(numericVector)
logicalVector <- c(TRUE, FALSE)
class(logicalVector)
```

Matrix & Data Frame
```{r echo = T, eval = T}
matrixVariable <- matrix(c(1:10), nrow = 2)
matrixVariable
class(matrixVariable)
dfVariable <- data.frame(x1 = 1:5, x2 = 6:10)
dfVariable
class(dfVariable)
```

Lists
```{r echo = T, eval = T, tidy = T}
listVariable <- list(x1 = c('a', 'b'), x2 = 1:4, x3 = matrix(c(1:6), nrow = 2))
listVariable
class(listVariable)
class(listVariable$x1)
class(listVariable$x2)
class(listVariable$x3)
```

Maths basics
```{r eval = T, echo = T}
5 + 2
11 - 3
(4+7)/9*2
14 %/% 3 + 1
8^(1/3) + 10%%6
```

Trigonometric functions
```{r eval = T, echo = T}
someArc <- 90*pi/180
sin(someArc)
atanValue <- atan(0.89)
atanValue*180/pi
```

Logical operations
```{r eval = T, echo = T}
5 >= 1
10%%2 == 0
!FALSE
5L | 11.1 <= 6
```

Using functions
```{r echo = T, eval = T}
tTestResult <- t.test(numericVector, integerVector)
print(tTestResult)
```

Addressing vector
```{r echo = T, eval = T}
addressVec <- seq(from = 1, to = 20, by = 2)
addressVec
addressVec[5]
addressVec[-5]
```

Wrong and correct addressing of vector
```{r echo = T, eval =T, error = T}
addressVec[1,5]
addressVec[-1,-5,-6]
addressVec[c(1,5)]
addressVec[-c(1,5,6)]
```

Conditional assignment to vector
```{r echo = T, eval =T, error = T}
addressVec[which(addressVec >= 6)] <- 'o..O'
addressVec
```

Data frame addressing
```{r echo = T, eval =T, error = T}
addressDF <- data.frame(C1 = seq(1,10), C2 = seq(11,20), C3 = seq(21,30))
addressDF[5, 2]
addressDF[6, ]
addressDF[c(6, 7), 1:3]
addressDF$C2[5]
addressDF$C3[c(6:9)]
```

List addressing
```{r echo = T, eval =T, error = T}
addressList <- list(x1 = c('a', 'b'), x2 = 1:4, x3 = matrix(c(1:6), nrow = 2))
addressList[[2]]
addressList$x2[3]
addressList[[3]]
addressList[[3]][, 3]
```

`if... else` statements

```{r}
#Simple
coinToss <- sample(c(0,1), 3, replace = T)
coinToss
ifelse(coinToss > 0, 'Win!', 'Looser!')

#Complex
coinToss <- sample(c(0,1), 3, replace = T)
sumOfCoins <- sum(coinToss)# we can have four states here 0,1,2,3
sumOfCoins

if (sumOfCoins == 0) {
  'Looser'
} else if (sumOfCoins == 1) {
  'Barely alive'
} else if (sumOfCoins >= 2) {
  'Big Winner!!!'
}

#Or because the functions realting to particular conditions are simple, you
#can make nested ifelse
ifelse(sum(sumOfCoins) == 0, 'Looser',
       ifelse(sum(sumOfCoins) == 1, 'Barely alive', 'Big Winner!!!'))
```

`for` and `while` loops
```{r}
exampleDF <- data.frame(norm = rnorm(100, 5, 1.5),
                        chisq = rchisq(100, 3),
                        pois = rpois(100, 3.33))
distroMeans <- vector('list', length(exampleDF)) #prealocate space

for (i in seq_along(exampleDF)) {
  distroMeans[[i]] <- summary(exampleDF[[i]])
}

distroMeans
```

```{r popPlot, fig.cap = 'Population dynamics', fig.height = 3.5}
#define initial parameters
N_0   <- 100
rf    <- 1.34
sf    <- 0.75
gf    <- rf*sf
n_gen <- 1:101
#we could repeat this inside loop, however we know the number of iterations
#thus calculating noise vector outside loop is better for performance
noise <- sample(-3:2, 100, replace = T)
t <- 1
geners <- vector('integer', length = length(n_gen))
geners[1] <- N_0
while (t < 101) {
  geners[c(t + 1)] <- floor((geners[t] * gf + noise[t]))
  t <- t + 1
}

plot(n_gen, geners, type = 'l',
     main = 'Population dynamics',
     xlab = 'generation', ylab = 'number of animals',
     col = 'blue', lty = 3)
```

Shortcut to loops AKA `apply` functions family
```{r}
testApply  <- list(x = 1:10, b = rep(c(4,5), times = 5), c = rpois(5, 25))
testLapply <- lapply(testApply, mean)
testSapply <- sapply(testApply, mean)

#Here we combine previous results with another lapply,
#and use function with multiple arguments.
lapply(testSapply, rpois, n = 15)
head(sapply(testSapply, rpois, n = 15))

testApplyFun <- 1:5
lapply(testApplyFun, function(x) x + 0.5)
sapply(testApplyFun, function(x) x + 0.5)
```

```{r lapplyPopPlot, fig.cap = 'Population dynamics with different starting point', fig.height = 3.5}
popDyn <- function(N_0, rf = 1.34, sf = 0.75, n_gen) {
  gf    <- rf*sf
  noise <- sample(-3:2, 100, replace = T)
  t <- 1
  geners <- vector('integer', length = n_gen)
  geners[1] <- N_0
  while (t < 101) {
  geners[c(t + 1)] <- floor((geners[t] * gf + noise[t]))
  t <- t + 1
  }
  return(geners)
}

N_0   <- c(50, 100, 150, 200, 250)
popSizes <- lapply(N_0, popDyn, n_gen = 101)

n_gen = 101
plot(1:n_gen, popSizes[[5]], type = 'l',
     main = 'Population dynamics',
     xlab = 'generation', ylab = 'number of animals',
     col = 'blue', lty = 3, ylim = c(-10, max(popSizes[[5]])))
lines(1:n_gen, popSizes[[4]], col = 'green')
lines(1:n_gen, popSizes[[3]], col = 'red')
lines(1:n_gen, popSizes[[2]], col = 'yellow')
lines(1:n_gen, popSizes[[1]], col = 'black')
```

Dice rolling as example of vectorization
```{r echo = T, eval =T, error = T}
niceDice <- seq(1,6)
rollDiceOne <- sample(niceDice, 1000, replace = T)
rollDiceTwo <- sample(niceDice, 1000, replace = T)
sumDiceRolls <- rollDiceOne + rollDiceTwo

#wrong sequence of events
cashDiceRolls <- sumDiceRolls
cashDiceRolls[which(cashDiceRolls %% 2 == 1 & cashDiceRolls <= 5)] <- 30
cashDiceRolls[which(cashDiceRolls %% 2 == 1 & cashDiceRolls > 5)] <- 50
cashDiceRolls[which(cashDiceRolls %% 2 == 0)] <- -70
sum(cashDiceRolls)

#correct sequence of events
cashDiceRolls <- sumDiceRolls
cashDiceRolls[which(cashDiceRolls %% 2 == 0)] <- 70
cashDiceRolls[which(cashDiceRolls %% 2 == 1 & cashDiceRolls > 5)] <- 50
cashDiceRolls[which(cashDiceRolls %% 2 == 1 & cashDiceRolls <= 5)] <- 30
cashDiceRolls[which(cashDiceRolls  == 70)] <- -70
sum(cashDiceRolls)
```

```{r echo = FALSE, cassino-plot, fig.cap = 'How to lose cash in casino', out.width='60%', fig.align= 'center'}
plot(seq(1,1000),cumsum(cashDiceRolls), type = 'l')
```

Distributions

Simple Beta example
```{r echo = T, eval =T, error = T}
rbeta(25, 10, 3)
```

```{r echo = FALSE, betaDensity-plot, fig.cap = 'Example of beta distribution', out.width='60%', fig.align= 'center'}
plot(seq(0, 1, length = 300), dbeta(seq(0, 1, length = 300), 8, 3), type = 'l')
```

Tidverse

```{r echo = TRUE, eval = TRUE}
wideTable <- data.frame(male = 1:10, female = 3:12)
wideTable
```

Converting wide to long table (and back)
```{r echo = TRUE, eval = TRUE, message = FALSE}
library('tidyr')
library('dplyr')
longTable <- gather(wideTable, key = sex, value = number)
longTable[7:12, ]
```

```{r echo = TRUE, eval = TRUE}
wideTable2 <- data.frame(male = 1:10,
                         female = 3:12, 
                         type = rep(c('bacteria', 'virus'), times = 5),
                         group = rep(c('a', 'b'), each = 5))
wideTable2
gather(wideTable2, sex, number, -c(3, 4))[7:13, ]
```

```{r echo = TRUE, eval = TRUE}
wideTable3 <- data.frame(male = rep(1:10, each = 2),
                         female = rep(3:12, times = 2), 
                         type = rep(c('bacteria','virus'), times = 10),
                         group = rep(c('a','b'), each = 10),
                         day = 1:10)
wideTable3[7:13, ]
gather(wideTable3, sex, number, 1:2) %>%
  spread(group, number) %>% .[7:13, ]
```

Separating column
```{r echo = TRUE, eval = TRUE}
wideTable4 <- data.frame(type = rep(c('bacteria.a', 'virus.a'), times = 5))
wideTable4
wideTable4 %>% separate(type, c('organism', 'type'), sep = '\\.')
```

Selecting and renaming column
```{r echo = TRUE, eval = TRUE}
head(select(wideTable3, starts_with('gr')), 3)
head(select(wideTable3, -starts_with('gr')), 3)
head(select(wideTable3, contains('ale')), 3)
head(rename(wideTable3, Male = male), 3)
```

Filtering rows
```{r echo = TRUE, eval = TRUE}
filter(wideTable3, group == 'a')
filter(wideTable3, type != 'bacteria')
filter(wideTable3, near(female, 5))
```

Mutating and transmuting (creating new values in (new)columns)
```{r echo = TRUE, eval = TRUE}
select(wideTable3, male) %>%
  mutate(cumulativeMaleSum = cumsum(male)) %>%
  head(5)

select(wideTable3, male, female) %>%
  mutate(cumulativeMaleSum = cumsum(male),
         femaleLog = log(female),
         cumSumFemaleLog = cumsum(femaleLog)) %>%
  head(5)
```

Grouping and summarizing
```{r echo = TRUE, eval = TRUE}
wideTable3 %>% 
  group_by(type) %>% 
  summarise(meanF = mean(female),
            meanM = mean(male),
            numberOfDays = (range(day)[2]-range(day)[1]),
            numberOfCases = n())
```